---
title: "FRANCHISE"
subtitle: "The Artometrics of Media Monopolies"
format:
  html:
    number-sections: false
    theme: none
    css: artometrics.css
    toc: true
    toc-depth: 2
    toc-title: "IN THIS REPORT"
    toc-location: left
    smooth-scroll: true
    embed-resources: true
    code-fold: true
    code-tools: true
    include-in-header: art-head.html
    include-before-body: art-body.html
editor:
  markdown:
    wrap: 72
---

```{r setup}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(scales)
library(glue)
library(plotly)

dir.create("charts", showWarnings = FALSE)
source("scripts/r/franchise_theme.R")
```

```{r load-data}
#| echo: false
#| warning: false
#| message: false

# TidyTuesday release used for every chart caption in this report.
tt_date <- FRANCHISE_DATA_DATE

media_franchises <- readr::read_csv(
  "data/media_franchises.csv",
  show_col_types = FALSE
)
```

```{r wrangle}
#| echo: false
#| warning: false
#| message: false

# ── Franchise-level summary tibble ─────────────────────────────────
franchises <- media_franchises %>%
  group_by(franchise, original_media, year_created, creators, owners) %>%
  summarize(
    categories    = length(revenue),
    total_revenue = sum(revenue),
    .groups       = "drop"
  )
```

```{r fast-facts-compute}
#| echo: false
#| warning: false
#| message: false
#| results: hide

# All values computed silently — hardcoded into Fast Facts HTML below.

n_franchises   <- n_distinct(media_franchises$franchise)         # 107
total_rev      <- sum(media_franchises$revenue)                  # ~572B
top_franchise  <- franchises %>% slice_max(total_revenue, n=1)  # Pokémon $91B
top_rev        <- top_franchise$total_revenue                    # 91.8

merch_share <- media_franchises %>%
  group_by(revenue_category) %>%
  summarise(rev = sum(revenue)) %>%
  mutate(pct = rev / sum(rev)) %>%
  filter(revenue_category == "Merchandise, Licensing & Retail") %>%
  pull(pct)                                                       # ~0.61

year_span <- max(media_franchises$year_created) -
             min(media_franchises$year_created)                  # 92
```

The global media industry runs on intellectual property — and a
surprisingly small number of franchises account for most of the money.
This report analyzes 107 properties from the TidyTuesday 2019 media
franchises dataset, each with at least \$4 billion in estimated lifetime
revenue as of mid-2019. Every franchise here has already cleared a
threshold that most IP never approaches. This is not a representative
sample of the media industry. It is a map of its ceiling.

Three charts, three angles on the same structural question: where does
franchise money actually come from, which properties generate it most
efficiently, and what does the ownership picture look like when you stop
treating subsidiaries as separate competitors? The answers are
Merchandise, Pokémon, and Disney — in that order.

## FAST FACTS

```{=html}
<div class="facts-grid">
  <div class="fact-box">
    <span class="fact-number">$91B</span>
    <span class="fact-label">Pokémon lifetime revenue — the highest-grossing media franchise ever recorded, more than the MCU, Star Wars, and Harry Potter put together</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">107</span>
    <span class="fact-label">Unique franchises in the dataset — every one of them has already cleared at least $4B in lifetime revenue</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">$4B+</span>
    <span class="fact-label">Minimum threshold for inclusion — most IP that has ever existed never gets close to this number</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">8</span>
    <span class="fact-label">Revenue categories tracked: Merchandise, Video Games, Box Office, Home Video, Comic/Manga, Music, TV, and Book Sales</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">92 yrs</span>
    <span class="fact-label">Span of the dataset — from Winnie the Pooh (1923) to Monster Strike (2015)</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">~61%</span>
    <span class="fact-label">Share of all tracked revenue that comes from Merchandise, Licensing & Retail — the dominant stream across the entire dataset</span>
  </div>
</div>
```

## DATASET CONTEXT

The TidyTuesday 2019 media franchises dataset covers 107 properties,
each with at least \$4B in estimated lifetime revenue as of mid-2019.
The figures are sourced from Wikipedia's list of highest-grossing media
franchises and compiled into eight revenue categories. Every franchise
here has already cleared a threshold that most IP never approaches —
this is not a representative sample of the media industry. It is a map
of its ceiling.

The \$4B minimum is not a small number. Most films, shows, games, and
books that have ever existed never come close to it. Being in this
dataset means a property has sustained commercial relevance across
multiple revenue streams, often across multiple decades. The franchises
below that line — culturally significant, critically acclaimed, beloved
— simply aren't here. What remains is the top tier: IP that didn't just
succeed but compounded.

Revenue here is a lifetime estimate, not an annual figure. A franchise
created in 1923 has had a century to accumulate; one created in 2013
has had six years. That asymmetry shapes every chart. Chart 1 shows the
raw totals and what they're made of. Chart 2 corrects for age by
normalizing revenue per year — a fairer fight. Chart 3 strips away the
ownership labels and asks what the leaderboard actually looks like when
subsidiaries are folded into their parent companies.

The `owners` column names the entity holding the IP rights as of 2019.
What it doesn't do is automatically consolidate subsidiaries under
parent companies. Disney owns Marvel Entertainment and 20th Century Fox,
but all three appear as separate rows in the raw data. Chart 3 exists
because of this gap — it asks what the ownership picture looks like when
you close it.

```{=html}
<div class="art-code-block">
  <details>
    <summary class="art-code-summary">
      <span class="art-lang-tag art-lang-sql">SQL</span>
      &nbsp;View dataset queries
    </summary>
    <pre class="art-code-pre" id="sql-block-1">-- MEDIA FRANCHISES: CLEANING AND FRANCHISE-LEVEL AGGREGATION
-- Mirrors the R cleaning logic in franchise.qmd
-- Source: TidyTuesday 2019-07-02 (Wikipedia via rfordatascience)

-- ── STEP 1: FRANCHISE-LEVEL SUMMARY ──────────────────────────────
-- Roll up from row-per-franchise-per-category to one row per franchise.

SELECT
    franchise,
    original_media,
    year_created,
    creators,
    owners,
    COUNT(revenue_category)         AS categories,
    ROUND(SUM(revenue), 2)          AS total_revenue,
    ROUND(2019 - year_created, 0)   AS years_active,
    ROUND(SUM(revenue)
          / NULLIF((2019 - year_created), 0), 4)
                                    AS revenue_per_year
FROM media_franchises
GROUP BY
    franchise, original_media, year_created, creators, owners
ORDER BY total_revenue DESC;


-- ── STEP 2: TOP 20 BY TOTAL REVENUE ──────────────────────────────

SELECT
    franchise,
    original_media,
    year_created,
    owners,
    ROUND(SUM(revenue), 2)   AS total_revenue
FROM media_franchises
GROUP BY franchise, original_media, year_created, owners
ORDER BY total_revenue DESC
LIMIT 20;


-- ── STEP 3: DISNEY CONSOLIDATION ─────────────────────────────────
-- Flag all Disney-owned entities and roll them into one parent bucket.

SELECT
    CASE
        WHEN owners LIKE '%Walt Disney%' THEN 'The Walt Disney Company (consolidated)'
        ELSE owners
    END                             AS parent_company,
    COUNT(DISTINCT franchise)       AS franchises_owned,
    ROUND(SUM(revenue), 2)          AS total_revenue
FROM media_franchises
GROUP BY parent_company
HAVING COUNT(DISTINCT franchise) >= 2
ORDER BY total_revenue DESC
LIMIT 12;</pre>
    <button class="art-copy-btn" onclick="copyBlock('sql-block-1', this)">Copy</button>
  </details>
</div>

<div class="art-code-block">
  <details>
    <summary class="art-code-summary">
      <span class="art-lang-tag art-lang-python">Python</span>
      &nbsp;View EDA notebook
    </summary>
    <pre class="art-code-pre" id="python-block-1"># Media Franchises — Python EDA
# Mirrors the EDA logic used to validate this dataset before R analysis.
# Display-only — same data, same URL, different tool.

import pandas as pd
import numpy as np
import requests
from io import StringIO

# ── Load data ─────────────────────────────────────────────────────
url = (
    "https://raw.githubusercontent.com/rfordatascience/tidytuesday/"
    "main/data/2019/2019-07-02/media_franchises.csv"
)
df = pd.read_csv(StringIO(requests.get(url).text))

print(f"Shape: {df.shape[0]:,} rows × {df.shape[1]} columns")
print(f"\nColumns: {list(df.columns)}")

# ── Missing values ────────────────────────────────────────────────
print("\n── MISSING VALUES ───────────────────────────────────────")
missing = df.isnull().sum()
print(missing[missing > 0] if missing.sum() > 0 else "None")

# ── Revenue summary ───────────────────────────────────────────────
print("\n── REVENUE SUMMARY (Billions USD) ───────────────────────")
print(df["revenue"].describe().round(2))

# ── Top 5 franchises by total revenue ────────────────────────────
print("\n── TOP 5 FRANCHISES BY TOTAL REVENUE ───────────────────")
top5 = (df.groupby("franchise")["revenue"]
          .sum()
          .sort_values(ascending=False)
          .head(5)
          .round(1))
print(top5.to_string())

# ── Revenue by category ───────────────────────────────────────────
print("\n── REVENUE BY CATEGORY ──────────────────────────────────")
by_cat = (df.groupby("revenue_category")["revenue"]
            .sum()
            .sort_values(ascending=False)
            .round(1))
print(by_cat.to_string())

# ── Revenue by original media ─────────────────────────────────────
print("\n── REVENUE BY ORIGINAL MEDIA ────────────────────────────")
by_media = (df.groupby("original_media")["revenue"]
              .agg(["sum", "count"])
              .sort_values("sum", ascending=False)
              .round(1))
by_media.columns = ["total_rev_B", "n_rows"]
print(by_media.to_string())

# ── Disney subsidiaries in ownership column ───────────────────────
print("\n── DISNEY-RELATED OWNERS ────────────────────────────────")
disney_rows = df[df["owners"].str.contains("Walt Disney", na=False)]
disney_summary = (disney_rows.groupby("owners")["revenue"]
                  .sum()
                  .sort_values(ascending=False)
                  .round(1))
print(disney_summary.to_string())
print(f"\nDisney consolidated total: ${disney_summary.sum():.1f}B")</pre>
    <button class="art-copy-btn" onclick="copyBlock('python-block-1', this)">Copy</button>
  </details>
</div>
```

## CHART 1 — TOP 20 FRANCHISES BY REVENUE

```{r chart1-stacked-bar}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8

chart1_date <- tt_date

top20 <- franchises %>%
  slice_max(total_revenue, n = 20)

plot_data <- media_franchises %>%
  semi_join(top20, by = "franchise") %>%
  mutate(
    franchise        = fct_reorder(franchise, revenue, sum),
    revenue_category = fct_relevel(revenue_category, rev(names(category_colors)))
  )

label_data <- plot_data %>%
  group_by(franchise) %>%
  summarize(total_revenue = sum(revenue), .groups = "drop") %>%
  mutate(
    franchise_num = as.numeric(franchise),
    label = paste0("$", round(total_revenue, 0), "B")
  )

p1 <- ggplot(plot_data, aes(franchise, revenue, fill = revenue_category)) +
  geom_hline(yintercept = 50, color = art_highlight, linetype = "dashed",
             linewidth = 0.5, alpha = 0.6) +
  geom_col(width = 0.9) +
  geom_text(
    data = label_data,
    aes(x = franchise, y = total_revenue, label = label),
    inherit.aes = FALSE,
    hjust = -0.05,
    size = 2.8,
    color = art_dark
  ) +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = category_colors) +
  coord_flip() +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_artometrics() +
  theme(panel.grid.major.y = element_blank()) +
  labs(
    title = "IN MEDIA, <span style='color:#C0392B'>**MERCHANDISE**</span> IS THE REAL PRODUCT",
    subtitle = "THE RED IN EVERY BAR TELLS THE SAME STORY — LICENSING & RETAIL DWARFS EVERY OTHER REVENUE STREAM",
    x = NULL,
    y = "Revenue (Billions USD)",
    fill = "Category",
    caption = franchise_caption(
      paste0("Chart concept: David Robinson, TidyTuesday Screencast 2019-07-22 (youtu.be/1xsbTs9-a50)")
    )
  )

p1_plotly <- ggplotly(
  p1 + theme(plot.margin = margin(t = 16, r = 120, b = 16, l = 16)),
  tooltip = c("y", "x", "fill")
) %>%
  plotly_franchise_layout(
    "IN MEDIA, <span style='color:#C0392B'><b>MERCHANDISE</b></span> IS THE REAL PRODUCT",
    "THE RED IN EVERY BAR TELLS THE SAME STORY — LICENSING & RETAIL DWARFS EVERY OTHER REVENUE STREAM"
  ) %>%
  add_bar_value_annotations(
    label_data %>% mutate(x = total_revenue, y = franchise_num),
    "x", "y", "label",
    x_pad = 1.03
  )

p1_plotly

save_franchise_chart(p1, p1_plotly, "chart1_top20_revenue")
```

Pokémon at \$91B is not just the top of this chart — it is a different
kind of number. The gap between Pokémon and second-place Hello Kitty
(\$80B) is itself larger than the total lifetime revenue of Batman,
Spider-Man, or Dragon Ball. This is what it looks like when a franchise
achieves structural escape velocity: merchandise revenue compounding
over decades, driven by a product line that expands into every available
category simultaneously — cards, video games, plush, clothing, themed
food, and an anime that has functioned primarily as a 25-year commercial
for the card game. Pokémon was not built to be a franchise. It was built
to be an economy.

The color logic of this chart is the argument. Merchandise, Licensing &
Retail (red) dominates the fill for nearly every bar in the top 20.
This is not a quirk of how Wikipedia compiles revenue — it is a
structural feature of how large-scale IP actually makes money. The
content — films, games, episodes — functions primarily as **marketing
infrastructure** for the real product, which is the license. The
entertainment generates demand. Merchandise captures it.

Three franchises break visibly from the red-dominant pattern. **Mario**
is mostly dark blue — Nintendo has spent decades resisting the licensing
temptation that Disney and Sanrio embraced fully, keeping the IP tightly
coupled to its own hardware ecosystem. **Shōnen Jump / Jump Comics** is
mostly purple (Comic or Manga) because its revenue structure is a
publishing business — magazines and tankōbon volumes, not T-shirts. And
**Star Wars** shows the most balanced mix: box office, merchandise, home
video, and games each contributing meaningfully. Star Wars is the model
of a fully diversified franchise — which is why Disney paid \$4B for
Lucasfilm in 2012.

```{=html}
<div class="art-code-block">
  <details>
    <summary class="art-code-summary">
      <span class="art-lang-tag art-lang-r">R</span>
      &nbsp;View Chart 1 code
    </summary>
    <pre class="art-code-pre" id="r-block-1">category_colors <- c(
  "Merchandise, Licensing & Retail" = art_highlight,
  "Video Games/Games"               = art_secondary,
  "Box Office"                      = "#E67E22",
  "Home Video/Entertainment"        = "#27AE60",
  "Comic or Manga"                  = "#8E44AD",
  "Music"                           = "#F39C12",
  "TV"                              = "#16A085",
  "Book sales"                      = "#2980B9"
)

top20 <- franchises %>% slice_max(total_revenue, n = 20)

plot_data <- media_franchises %>%
  semi_join(top20, by = "franchise") %>%
  mutate(
    franchise        = fct_reorder(franchise, revenue, sum),
    revenue_category = fct_reorder(revenue_category, revenue, sum)
  )

label_data <- plot_data %>%
  group_by(franchise) %>%
  summarize(total_revenue = sum(revenue), .groups = "drop")

p1 <- ggplot(plot_data, aes(franchise, revenue, fill = revenue_category)) +
  geom_hline(yintercept = 50, color = art_highlight, linetype = "dashed",
             linewidth = 0.5, alpha = 0.6) +
  geom_col() +
  geom_text(
    data = label_data,
    aes(x = franchise, y = total_revenue,
        label = paste0("$", round(total_revenue, 0), "B")),
    inherit.aes = FALSE, hjust = -0.1, size = 2.8, color = art_dark
  ) +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = category_colors) +
  coord_flip() +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_artometrics() +
  theme(panel.grid.major.y = element_blank()) +
  labs(
    title   = "IN MEDIA, &lt;span style='color:#C0392B'&gt;**MERCHANDISE**&lt;/span&gt; IS THE REAL PRODUCT",
    subtitle = "THE RED IN EVERY BAR TELLS THE SAME STORY — LICENSING & RETAIL DWARFS EVERY OTHER REVENUE STREAM",
    x = NULL, y = "Revenue (Billions USD)", fill = "Category",
    caption = "Data: TidyTuesday 2019-07-02\n— ARTOMETRICS"
  )

ggsave("chart1_top20_revenue.png", plot = p1,
       path = "charts", width = 12, height = 7, dpi = 300, bg = "white")</pre>
    <button class="art-copy-btn" onclick="copyBlock('r-block-1', this)">Copy</button>
  </details>
</div>
```

## CHART 2 — REVENUE PER YEAR OF EXISTENCE

```{r chart2-lollipop}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8

chart2_date <- tt_date

chart2_data <- franchises %>%
  mutate(
    years_active     = 2019 - year_created,
    revenue_per_year = total_revenue / years_active,
    is_top           = as.character(franchise == "Pokémon")
  ) %>%
  slice_max(revenue_per_year, n = 15) %>%
  mutate(
    franchise = fct_reorder(franchise, revenue_per_year),
    franchise_num = as.numeric(franchise),
    label = paste0("$", round(revenue_per_year, 1), "B/yr")
  )

p2 <- ggplot(chart2_data, aes(x = franchise, y = revenue_per_year)) +
  geom_hline(yintercept = 2, color = art_highlight, linetype = "dashed",
             linewidth = 0.5, alpha = 0.6) +
  geom_segment(
    aes(xend = franchise, y = 0, yend = revenue_per_year, color = is_top),
    linewidth = 0.8
  ) +
  geom_point(aes(color = is_top, size = is_top)) +
  geom_text(aes(label = label), hjust = -0.2, size = 3, color = art_dark) +
  scale_color_manual(values = c("TRUE" = art_highlight, "FALSE" = art_secondary),
                     guide = "none") +
  scale_size_manual(values = c("TRUE" = 7, "FALSE" = 4.5), guide = "none") +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.28))) +
  coord_flip() +
  theme_artometrics() +
  theme(panel.grid.major.y = element_blank()) +
  labs(
    title = "<span style='color:#C0392B'>POKÉMON</span> EARNS MORE PER YEAR THAN MOST FRANCHISES EARN IN A LIFETIME",
    subtitle = "~$4B ANNUALLY SINCE 1996 — MORE THAN HELLO KITTY, WHICH IS TWICE ITS AGE",
    x = NULL,
    y = "Revenue Per Year (Billions USD)",
    caption = franchise_caption("Revenue ÷ (2019 − year created)")
  )

p2_plotly <- ggplotly(
  p2 + theme(plot.margin = margin(t = 16, r = 120, b = 16, l = 16)),
  tooltip = c("y", "x")
) %>%
  plotly_franchise_layout(
    "<span style='color:#C0392B'><b>POKÉMON</b></span> EARNS MORE PER YEAR THAN MOST FRANCHISES EARN IN A LIFETIME",
    "~$4B ANNUALLY SINCE 1996 — MORE THAN HELLO KITTY, WHICH IS TWICE ITS AGE"
  ) %>%
  add_bar_value_annotations(
    chart2_data %>% mutate(x = revenue_per_year, y = franchise_num),
    "x", "y", "label",
    x_pad = 1.05
  )

p2_plotly

save_franchise_chart(p2, p2_plotly, "chart2_revenue_per_year")
```

Total revenue rewards longevity. Revenue per year rewards efficiency.
Normalizing by age puts a 1928 franchise and a 1996 franchise on
comparable footing — and the resulting ranking surfaces something the
raw totals hide. Pokémon's \$4B/yr is not explained by any single hit.
It is explained by a system: the video games create new entry points
every hardware generation; the trading card game creates perpetual
demand for new product; the anime provides continuous character exposure
that drives card and game sales. Each component feeds the others in a
loop that doesn't require any single cultural moment to sustain. Pokémon
doesn't just earn well — it earns continuously, from multiple
directions, at the same time.

Several franchises that ranked very high on total revenue drop here when
normalized by time. **Hello Kitty** (\$80B total, second overall) falls
to fourth at \$1.8B/yr — because it has been generating that revenue
since 1974, making 45 years the denominator that brings the rate down.
**Winnie the Pooh** and **Mickey Mouse** drop further still. Their
massive lifetime totals are the product of nearly a century of
operation. The per-year lens is unforgiving to age: a franchise that
earns \$1B/yr for 90 years scores lower on this chart than one earning
\$2B/yr for 25.

The franchises that rise on this chart share a structural trait: they
earn across multiple simultaneous streams rather than sequentially. A
film franchise earns at release, then waits for the next film. A
trading-card-plus-video-game-plus-anime ecosystem earns every month,
every quarter, at every retail touchpoint. The per-year chart is really
a chart of business model design.

```{=html}
<div class="art-code-block">
  <details>
    <summary class="art-code-summary">
      <span class="art-lang-tag art-lang-r">R</span>
      &nbsp;View Chart 2 code
    </summary>
    <pre class="art-code-pre" id="r-block-2">chart2_data <- franchises %>%
  mutate(
    years_active     = 2019 - year_created,
    revenue_per_year = total_revenue / years_active,
    is_top           = as.character(franchise == "Pokémon")
  ) %>%
  slice_max(revenue_per_year, n = 15) %>%
  mutate(franchise = fct_reorder(franchise, revenue_per_year))

p2 <- ggplot(chart2_data, aes(x = franchise, y = revenue_per_year)) +
  geom_hline(yintercept = 2, color = art_highlight, linetype = "dashed",
             linewidth = 0.5, alpha = 0.6) +
  geom_segment(
    aes(xend = franchise, y = 0, yend = revenue_per_year, color = is_top),
    linewidth = 0.8
  ) +
  geom_point(aes(color = is_top, size = is_top)) +
  geom_text(
    aes(label = paste0("$", round(revenue_per_year, 1), "B/yr")),
    hjust = -0.2, size = 3, color = art_dark
  ) +
  scale_color_manual(values = c("TRUE" = art_highlight, "FALSE" = art_secondary),
                     guide = "none") +
  scale_size_manual(values = c("TRUE" = 7, "FALSE" = 4.5), guide = "none") +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.28))) +
  coord_flip() +
  theme_artometrics() +
  theme(panel.grid.major.y = element_blank()) +
  labs(
    title    = "&lt;span style='color:#C0392B'&gt;POKÉMON&lt;/span&gt; EARNS MORE PER YEAR THAN MOST FRANCHISES EARN IN A LIFETIME",
    subtitle = "~$4B ANNUALLY SINCE 1996 — MORE THAN HELLO KITTY, WHICH IS TWICE ITS AGE",
    x = NULL, y = "Revenue Per Year (Billions USD)",
    caption = "Data: TidyTuesday 2019-07-02 | Revenue ÷ (2019 − year created)\n— ARTOMETRICS"
  )

ggsave("chart2_revenue_per_year.png", plot = p2,
       path = "charts", width = 12, height = 7, dpi = 300, bg = "white")</pre>
    <button class="art-copy-btn" onclick="copyBlock('r-block-2', this)">Copy</button>
  </details>
</div>
```

## CHART 3 — THE DISNEY EMPIRE, CONSOLIDATED

```{r chart3-disney-non-consolidated}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8

chart3a_date <- tt_date

non_consolidated <- media_franchises |>
  group_by(owners) |>
  summarise(
    total_revenue    = sum(revenue, na.rm = TRUE),
    franchises_owned = n_distinct(franchise),
    .groups          = "drop"
  ) |>
  filter(franchises_owned >= 2) |>
  slice_max(total_revenue, n = 14) |>
  mutate(
    owners      = fct_reorder(owners, total_revenue),
    owners_num  = as.numeric(owners),
    disney_flag = case_when(
      owners == "The Walt Disney Company"                        ~ "disney_direct",
      owners == "Marvel Entertainment (The Walt Disney Company)" ~ "disney_marvel",
      owners == "20th Century Fox (The Walt Disney Company)"     ~ "disney_fox",
      TRUE                                                       ~ "other"
    ),
    label = paste0("$", round(total_revenue), "B")
  )

p3a <- ggplot(non_consolidated,
       aes(x = owners, y = total_revenue, fill = disney_flag)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = label), hjust = -0.12, color = art_dark, size = 3.2) +
  scale_fill_manual(values = c(
    "disney_direct" = art_highlight,
    "disney_marvel" = "#D4695E",
    "disney_fox"    = "#E8A09A",
    "other"         = art_secondary
  )) +
  scale_y_continuous(
    labels = dollar_format(suffix = "B"),
    expand = expansion(mult = c(0, 0.22))
  ) +
  coord_flip() +
  theme_artometrics() +
  theme(axis.text.y = element_text(size = 9)) +
  labs(
    title    = "THREE OF THESE BARS BELONG TO <span style='color:#C0392B'>**THE SAME COMPANY**</span>",
    subtitle = "MARVEL ENTERTAINMENT AND 20TH CENTURY FOX APPEAR AS SEPARATE COMPETITORS — BOTH ARE WHOLLY OWNED BY DISNEY",
    x        = NULL,
    y        = "Total Revenue (Billions USD)",
    caption  = franchise_caption("Wikipedia via TidyTuesday")
  )

p3a_plotly <- ggplotly(
  p3a + theme(plot.margin = margin(t = 16, r = 120, b = 16, l = 16)),
  tooltip = c("y", "x", "fill")
) %>%
  plotly_franchise_layout(
    "THREE OF THESE BARS BELONG TO <span style='color:#C0392B'><b>THE SAME COMPANY</b></span>",
    "MARVEL ENTERTAINMENT AND 20TH CENTURY FOX APPEAR AS SEPARATE COMPETITORS — BOTH ARE WHOLLY OWNED BY DISNEY"
  ) %>%
  add_bar_value_annotations(
    non_consolidated %>% mutate(x = total_revenue, y = owners_num),
    "x", "y", "label",
    x_pad = 1.03
  )

p3a_plotly

save_franchise_chart(p3a, p3a_plotly, "chart3a_disney_non_consolidated")
```

The raw data lists "The Walt Disney Company," "Marvel Entertainment (The
Walt Disney Company)," and "20th Century Fox (The Walt Disney Company)"
as three separate owners. Technically accurate. Strategically misleading.
Marvel was acquired in 2009 for \$4B. Fox was acquired in 2019 for \$71B.
Both are now wholly owned subsidiaries. Their revenue goes to the same
place. This is how corporate IP ownership actually works. You don't buy
franchises — you buy the companies that own franchises, and those
companies keep operating under their own names. Marvel Studios still
greenlit its own films. Fox kept its lot in Century City. But the
revenue goes to Burbank.

```{r chart3-disney-consolidated}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8

chart3b_date <- tt_date

disney_consolidated <- media_franchises |>
  mutate(
    parent_company = case_when(
      str_detect(owners, "Walt Disney") ~ "The Walt Disney Company\n(consolidated)",
      TRUE ~ owners
    )
  ) |>
  group_by(parent_company) |>
  summarise(
    total_revenue    = sum(revenue, na.rm = TRUE),
    franchises_owned = n_distinct(franchise),
    .groups          = "drop"
  ) |>
  filter(franchises_owned >= 2) |>
  slice_max(total_revenue, n = 12) |>
  mutate(
    parent_company = fct_reorder(parent_company, total_revenue),
    parent_num     = as.numeric(parent_company),
    is_disney      = str_detect(parent_company, "Walt Disney"),
    label          = paste0("$", round(total_revenue), "B")
  )

p3b <- ggplot(disney_consolidated,
       aes(x = parent_company, y = total_revenue, fill = is_disney)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = label), hjust = -0.12, color = art_dark, size = 3.2) +
  scale_fill_manual(
    values = c("FALSE" = art_secondary, "TRUE" = art_highlight)
  ) +
  scale_y_continuous(
    labels = dollar_format(suffix = "B"),
    expand = expansion(mult = c(0, 0.22))
  ) +
  coord_flip() +
  theme_artometrics() +
  theme(axis.text.y = element_text(size = 9)) +
  labs(
    title    = "<span style='color:#C0392B'>**DISNEY**</span> DOESN'T JUST OWN THE MOST IP — IT OWNS THE COMPANIES THAT OWN THE IP",
    subtitle = "CONSOLIDATED BY PARENT COMPANY — MARVEL ENTERTAINMENT AND 20TH CENTURY FOX FOLDED INTO DISNEY'S TRUE TOTAL",
    x        = NULL,
    y        = "Total Revenue (Billions USD)",
    caption  = franchise_caption("Wikipedia via TidyTuesday")
  )

p3b_plotly <- ggplotly(
  p3b + theme(plot.margin = margin(t = 16, r = 120, b = 16, l = 16)),
  tooltip = c("y", "x", "fill")
) %>%
  plotly_franchise_layout(
    "<span style='color:#C0392B'><b>DISNEY</b></span> DOESN'T JUST OWN THE MOST IP — IT OWNS THE COMPANIES THAT OWN THE IP",
    "CONSOLIDATED BY PARENT COMPANY — MARVEL ENTERTAINMENT AND 20TH CENTURY FOX FOLDED INTO DISNEY'S TRUE TOTAL"
  ) %>%
  add_bar_value_annotations(
    disney_consolidated %>% mutate(x = total_revenue, y = parent_num),
    "x", "y", "label",
    x_pad = 1.03
  )

p3b_plotly

save_franchise_chart(p3b, p3b_plotly, "chart3b_disney_consolidated")
```

Add them up and Disney's true total is **\$426 billion**. Nintendo at
\$49B — still second — now represents about 11 cents on Disney's dollar.
But here's what the math actually surfaces: Marvel (\$20B) plus Fox
(\$17B) should consolidate to \$296B. The remaining **\$130 billion
comes from Disney-owned entities too small to crack the top 14
individually** — Pixar, Lucasfilm, ABC, ESPN. Subsidiaries whose
franchise rows exist in the dataset but don't rank highly enough to
appear as standalone bars. Individually invisible. Collectively, they
add up to more than Nintendo's entire lifetime output.

That is the actual scale of the Disney empire. Not one dominant company.
Not even three. An ecosystem of wholly-owned studios, publishers, and
licensors — some famous, some not — each running their own operations,
all consolidating upward to the same balance sheet. The chart makes
visible something the raw data quietly obscures: **in IP, ownership
structure is as strategically important as the franchises themselves**.
Disney didn't just build great IP. It bought the companies that owned
great IP, let them keep their identities, and harvested everything
upward.

```{=html}
<div class="art-code-block">
  <details>
    <summary class="art-code-summary">
      <span class="art-lang-tag art-lang-r">R</span>
      &nbsp;View Chart 3 code
    </summary>
    <pre class="art-code-pre" id="r-block-3">non_consolidated <- media_franchises |>
  group_by(owners) |>
  summarise(total_revenue = sum(revenue, na.rm = TRUE),
            franchises_owned = n_distinct(franchise), .groups = "drop") |>
  filter(franchises_owned >= 2) |>
  slice_max(total_revenue, n = 14) |>
  mutate(owners = fct_reorder(owners, total_revenue),
         disney_flag = case_when(
           owners == "The Walt Disney Company"                        ~ "disney_direct",
           owners == "Marvel Entertainment (The Walt Disney Company)" ~ "disney_marvel",
           owners == "20th Century Fox (The Walt Disney Company)"     ~ "disney_fox",
           TRUE                                                       ~ "other"
         ))

disney_consolidated <- media_franchises |>
  mutate(parent_company = case_when(
    str_detect(owners, "Walt Disney") ~ "The Walt Disney Company\n(consolidated)",
    TRUE ~ owners
  )) |>
  group_by(parent_company) |>
  summarise(total_revenue = sum(revenue, na.rm = TRUE),
            franchises_owned = n_distinct(franchise), .groups = "drop") |>
  filter(franchises_owned >= 2) |>
  slice_max(total_revenue, n = 12) |>
  mutate(parent_company = fct_reorder(parent_company, total_revenue),
         is_disney = str_detect(parent_company, "Walt Disney"))

# Chart 3a — Non-consolidated
ggsave("chart3a_disney_non_consolidated.png", path = "charts",
       width = 12, height = 7, dpi = 300, bg = "white")

# Chart 3b — Consolidated
ggsave("chart3b_disney_consolidated.png", path = "charts",
       width = 12, height = 7, dpi = 300, bg = "white")</pre>
    <button class="art-copy-btn" onclick="copyBlock('r-block-3', this)">Copy</button>
  </details>
</div>
```

## LIMITATIONS

This dataset reflects revenue estimates compiled from Wikipedia as of
mid-2019, and carries several important caveats. Revenue figures are
sourced inconsistently — some represent lifetime totals through 2019,
others may reflect different windows or methodologies. Franchises are
only included if they surpassed an estimated \$4B threshold, meaning
smaller but culturally significant IP is absent entirely. The ownership
and creator fields reflect the state of the industry at the time of
collection and do not account for subsequent acquisitions. Revenue
categories were consolidated from over 60 raw Wikipedia subcategories
into 8 groups, which involves judgment calls documented in the original
TidyTuesday cleaning script.

The "Merchandise, Licensing & Retail" category is particularly broad
and may include revenue streams that other analyses would track
separately. The dominance of merchandise in Chart 1 is structurally
real, but the exact percentage share depends on how individual Wikipedia
editors categorized edge cases — which is not fully auditable from this
dataset alone. All revenue figures should be treated as
order-of-magnitude approximations rather than audited financial data.

Chart 2 divides total lifetime revenue by years of existence as of 2019.
This is a simple normalization, not a modeled annual revenue figure.
A franchise with uneven revenue across its lifespan — a massive recent
hit, or decades of dormancy — will produce a per-year figure that doesn't
reflect any actual year's performance. The metric rewards consistent
earners and penalizes anything that earned most of its revenue in a
concentrated burst.

## CONCLUSION

The media franchise landscape is not flat — it is sharply skewed toward
a small number of IP that have mastered merchandise and sustained it
across decades. Pokémon stands alone at the top not because of box
office dominance but because it turned a video game into the world's
most successful licensing machine while simultaneously running an anime,
a card game, and a global trading ecosystem. The Artometrics takeaway
from this dataset is a single structural insight: **the most valuable
thing a franchise can do is give people something to own, collect, and
trade — not just something to watch**.

Disney's story is a different kind of lesson. The company didn't just
build great IP — it systematically acquired the companies that owned
great IP and rolled them into a single consolidated empire. The \$130B
gap between Disney's apparent total and its true total is a number that
only becomes visible when you collapse the ownership structure the raw
data obscures. That gap, by itself, is larger than Nintendo's entire
lifetime output. In IP, what you own is one thing. What your
subsidiaries own is another. Disney understood the difference earlier
and more completely than anyone.

The two insights together describe the same underlying dynamic from
different angles. Merchandise is how IP extracts value from audiences.
Acquisition is how corporations extract value from IP. The companies
that have mastered both — Disney chief among them — are not competing
in the same game as everyone else. They are operating at a different
level of the stack entirely.

## REFERENCES

```{=html}
<div class="art-references">
  <div class="art-ref-item">
    TidyTuesday (2019-07-02). <em>Media Franchise Revenues</em>.
    R for Data Science Community.
    <a href="https://github.com/rfordatascience/tidytuesday/tree/main/data/2019/2019-07-02"
       target="_blank">
      github.com/rfordatascience/tidytuesday/…/2019-07-02
    </a>
  </div>
  <div class="art-ref-item">
    Wikipedia. <em>List of Highest-Grossing Media Franchises</em>.
    Retrieved via TidyTuesday cleaning script (revenue.R).
  </div>
  <div class="art-ref-item">
    Robinson, D. (2019, July 22). <em>Analyzing Franchise Revenue —
    TidyTuesday Screencast</em>. YouTube.
    <a href="https://youtu.be/1xsbTs9-a50" target="_blank">
      youtu.be/1xsbTs9-a50
    </a>
    — Chart 1 concept adapted from this screencast with original
    Artometrics styling applied.
  </div>
</div>
```

## EDITOR'S NOTE

```{=html}
<p class="art-p">
This report was researched, written, designed, and produced in active
collaboration with Claude AI (Anthropic). The data pipeline, statistical
analysis, chart design, written analysis, narrative structure, and visual
styling were all developed through a directed partnership between human
editorial judgment and AI execution.
</p>
<p class="art-p">
Artometrics was built on the premise that rigorous analysis and honest
process are not in conflict. The research questions, editorial instincts,
interpretive framing, and brand vision are ours. The execution — every
line of R code, every paragraph of analysis, every design decision — was
a collaboration. We document this not as a disclaimer but as a
description of how we actually work, and as a position: we believe this
is what serious data journalism looks like when the tools available are
used honestly and at full capacity.
</p>
<p class="art-p">— Artometrics Editorial</p>
<p style="text-align:center; margin: 2rem 0 1rem;">
  <a href="https://github.com/Artometrics/franchise"
     target="_blank" rel="noopener"
     style="display:inline-block; font-family:'DM Sans',sans-serif;
            font-size:0.85rem; font-weight:600; letter-spacing:1px;
            text-transform:uppercase; color:#FFFFFF; background:#1C1C1E;
            border:2px solid #1C1C1E; border-radius:3px;
            padding:10px 24px; text-decoration:none;
            transition:background 0.15s, color 0.15s;">
    View on GitHub
  </a>
</p>
```